2D simulated case study: Diagnosis of the HSGP approximation

### 2022-05-15 Paper/Case-study_2D&3D-Simulated-data/2Ddataset/Rnb_case2D_diagnosis.Rmd
reorganize files

```{r message=FALSE, warn=FALSE, eval=TRUE}
library(rstan)
library(cmdstanr)
options(mc.cores = parallel::detectCores())
rstan_options(auto_write = FALSE)
library(posterior)
library(bayesplot)
library(RColorBrewer)
library(ggplot2)
library(geoR)
```

@@

where $k(\boldsymbol{x}, \boldsymbol{x}', \theta)$ is a squared exponential kernel as a function of input values $\boldsymbol{x}$ and parameters $\theta=(\alpha,\ell)$, where $\alpha$ is the marginal variance of the GP and $\ell$ the lengthscale of the kernel.

@@

Matrix of 2D-predictive input points `x_pred`, length of `x_pred` (`N_pred`)

```{r eval=TRUE}
x_pred <- pred_grid(coords= c(-1,1), y.coords= c(-1,1), by= 0.04)  
N_pred <- dim(x_pred)[1]
N_pred
```

@@

Half range of the input domain

```{r eval=TRUE, message=FALSE, warn=FALSE}
S <- abs((max(x_pred) - min(x_pred))/2)
S
```

@@

Marginal variance `gpscale` ($\alpha$) and lengthscale `lscale` ($\ell$)

@@

```{r eval=TRUE, results=FALSE}
simuGP_mod <- cmdstanr::cmdstan_model(stan_file = "stancode_SIMU_2D.stan")
fit_SIMU <- simuGP_mod$sample(data= standata,
                                iter_warmup=1,
                                iter_sampling=1,
                                chains=1, thin=1, init=0.5,
                                fixed_param=TRUE)
fit_SIMU <- read_stan_csv(fit_SIMU$output_files())
```

@@

```{r eval=TRUE, echo=FALSE}
f_true <- extract(fit_SIMU, pars="f_true", permuted = TRUE, inc_warmup = FALSE, include = TRUE)$f_true[1,]
xx <- unique(x_pred[,1])

@@

## Posterior predictive distributions

```{r eval=TRUE, include=TRUE, echo=FALSE}
gp12_f_pred <- summary(fit_GP, pars = c("f_pred"), probs = c(0.025, 0.5, 0.975), digits_summary = 4)$summary[,c(1,4,6)]
gp12_elpd <- summary(fit_GP, pars= c("elpd"), probs = c(0.025, 0.5, 0.975), digits_summary= 4)$summary[,c(1,4,6)]
gp12_time <- sum(get_elapsed_time(fit_GP)[2])/(fit_GP@sim$iter - fit_GP@sim$warmup)
gp12_mse <- sqrt(sum((f_true - gp12_f_pred[,1])^2)/N_pred)

col1 <- brewer.pal(11,"BrBG")[11:1] 
# brk <- seq(min(gp12_f_pred[,1]),max(gp12_f_pred[,1]), length.out= 12)

filled.contour(xx, xx, matrix(gp12_f_pred[,1],length(xx),length(xx)), levels = brk, col = brewer.pal(11,"BrBG")[11:1], plot.axes= {axis(1, tick=FALSE, line=-1.1, font=5, mgp= c(2.5, 0.1, 0), las=1, cex.axis=1.5); title(xlab =expression(x[1]), mgp= c(0.8, 1, 0), cex.lab=2, las=1, family="serif"); axis(2, tick=FALSE, line=0.4, font=5, mgp= c(2.5, 0.0, 0), las=1, cex.axis=1.5); points(x_pred[vv_sample,], cex=1, lwd=1, pch=1);  abline(h= seq(-1,1,0.5), col=grey(0.7)); lines(c(-1,-1,-0.5,-0.5,0,0,0.5,0.5,1,1), c(-1.03,1.03,-1.03,1.03,-1.03,1.03,-1.03,1.03,-1.03,1.03), col=grey(0.7), type="h")}, asp= 1, frame.plot=FALSE, xlim=c(-1.03,1.03), ylim=c(-1.03,1.03)) #; grid(col=grey(0.90), lty=1)
text(-0.3,0.9,"(b)   GP", font=2, cex=1.5);
text( par("usr")[1] - 0.30, mean(c(-1,1)), labels = expression(x[2]), srt = 0, xpd = TRUE, cex=2, font=1, family="serif")
```

@@

```{r eval=TRUE, include=TRUE, echo=FALSE}
error <- f_true - gp12_f_pred[,1]

col3 <- c(colorRampPalette(c("blue", "white"), bias=10)(6), colorRampPalette(c("white", "red"), bias=10)(6)[2:6])#brewer.pal(11,"RdBu")[1:11]
brk_error <- seq(min(error),max(error), length.out= 12)

@@

```{r eval=TRUE, include=TRUE}
m_QE <- function(c,l,S) ceiling(1.75 * c / (l/S))
l_QE <- function(c,m,S) round(S * 1.75 * c / m, 3)   
```

@@

as a function of $\ell$ and $S$, where $S$ is the half range of the input domain,

```{r eval=TRUE, include=TRUE}
c_vs_l_QE <- function(l,S){
    c =  3.2*l/S
    if(c < 1.2)
      c = 1.2
  c
}
```

# Diagnostic function over the lengthscale of the HSGP model

The lengthscale diagnosis function is defined as:

@@

```{r eval=TRUE, include=TRUE}
diagnostic <- function(l,l_hat) l_hat + 0.01 >= l
```

@@

```{r eval=TRUE, include=TRUE, results='hold'}
i <- 1
print(paste("Iteration i =", i), quote = FALSE)
```

@@

```{r eval=TRUE, include=TRUE}
if(i==1){
  l1[i] <- 0.5
  c1[i] <- c_vs_l_QE(l=l1[i], S)
  m1[i] <- m_QE(c=c1[i], l=l1[i], S=S)

  l2[i] <- 1
  c2[i] <- c_vs_l_QE(l=l2[i], S)
  m2[i] <- m_QE(c=c2[i], l=l2[i], S=S)
}
```

@@

```{r eval=TRUE, message=FALSE, warn=FALSE}
standata[[i]] <- list(D= D, 
                      M= c(m1[i], m2[i]),
                      M_nD= m1[i]*m2[i], 
                      indices= indices,
                      L= c(c1[i], c2[i]) * S, 
                      x_pred= x_pred, 
                      y_pred= y_pred, 
                      N_sample= N_sample, 
                      N_pred= N_pred, 
                      vv_sample= vv_sample
)
str(standata[[i]])
```

@@

```{r eval=TRUE, include=TRUE}
f <- summary(fit[[i]], pars = c("f"), probs = c(0.025, 0.5, 0.975), digits_summary = 4)$summary[,c(1,4,6)]

# rmse against data
residual <- standata[[i]]$y_pred[vv_sample] - f[,1]
rmse[i] <- round(sqrt(mean(residual^2)), 2)

# rmse against data
residual_gp <- gp12_f_pred[vv_sample] - f[,1]
rmse_gp[i] <- round(sqrt(mean(residual_gp^2)), 2)

@@

```{r eval=TRUE, include=TRUE}
l_hat1[i] <- round(summary(fit[[i]], pars = "lscale[1]", probs = c(0.05, 0.95))$summary[,1], 2)
l_hat2[i] <- round(summary(fit[[i]], pars = "lscale[2]", probs = c(0.05, 0.95))$summary[,1], 2)
```

@@

```{r eval=TRUE, include=TRUE}
check1[i] <- diagnostic(l1[i], l_hat1[i])
check2[i] <- diagnostic(l2[i], l_hat2[i])
```

@@

```{r eval=TRUE, include=TRUE}
diagnosis[[i]] <- data.frame(iter= c(i, i),
                              Dim= c("D1", "D2"),
                              l= c(l1[i], l2[i]),
                              c= c(c1[i], c2[i]),
                              m= c(m1[i], m2[i]),
                              l_hat= c(l_hat1[i], l_hat2[i]),
                              'l_hat > l' = c(check1[i], check2[i]),
                              rsme_gp = c(NA, rmse_gp[i]),
                              rsme = c(NA, rmse[i]),
                              R2 = c(NA, eR[i]),
                              elpd = c(NA, elpd[i]))
names(diagnosis[[i]]) <- c("iter", "Dim", "l", "c", "m", "l_hat", "l_hat > l", "rmse_gp", "rmse", "R2", "elpd")

@@

```{r eval=TRUE, include=TRUE, results='hold'}
i <- 2
print(paste("Iteration i =", i), quote = FALSE)
```

@@

```{r eval=TRUE, include=TRUE, results='hold'}
if(diagnostic(l1[i-1], l_hat1[i-1]) & i==2){
  m1[i] <- m1[i-1] + 2
  c1[i] <- c_vs_l_QE(l=l_hat1[i-1], S=S)
  l1[i] <- l_QE(c=c1[i], m=m1[i], S=S)

  m2[i] <- m2[i-1] + 2
  c2[i] <- c_vs_l_QE(l=l_hat2[i-1], S=S)
  l2[i] <- l_QE(c=c2[i], m=m2[i], S=S)
}
```

@@

```{r eval=TRUE, include=TRUE, results='hold'}
if(!diagnostic(l1[i-1], l_hat1[i-1])){
  l1[i] <- l_hat1[i-1]
  c1[i] <- c_vs_l_QE(l=l1[i], S=S)
  m1[i] <- m_QE(c=c1[i], l=l1[i], S=S)

  l2[i] <- l_hat2[i-1]
  c2[i] <- c_vs_l_QE(l=l2[i], S=S)
  m2[i] <- m_QE(c=c2[i], l=l2[i], S=S)
}
```

@@

```{r eval=TRUE}
standata[[i]] <- list(D= D, 
                      M= c(m1[i], m2[i]),
                      M_nD= m1[i]*m2[i], 
                      indices= indices,
                      L= c(c1[i], c2[i]) * S, 
                      x_pred= x_pred, 
                      y_pred= y_pred, 
                      N_sample= N_sample, 
                      N_pred= N_pred, 
                      vv_sample= vv_sample
)
str(standata[[i]])
```

@@

```{r eval=TRUE, include=TRUE}
l_hat1[i] <- round(summary(fit[[i]], pars = "lscale[1]", probs = c(0.05, 0.95))$summary[,1], 3)
l_hat2[i] <- round(summary(fit[[i]], pars = "lscale[2]", probs = c(0.05, 0.95))$summary[,1], 3)
```

@@

```{r eval=TRUE, include=TRUE, results='hold'}
i <- 3
print(paste("Iteration i =", i), quote = FALSE)
```

@@

```{r eval=TRUE, include=TRUE, results='hold'}
if(diagnostic(l1[i-1], l_hat1[i-1]) & i>2 & !diagnostic(l1[i-2], l_hat1[i-2])){
  m1[i] <- m1[i-1] + 2
  c1[i] <- c_vs_l_QE(l=l_hat1[i-1], S=S)
  l1[i] <- l_QE(c=c1[i], m=m1[i], S=S)

  m2[i] <- m2[i-1] + 2
  c2[i] <- c_vs_l_QE(l=l_hat2[i-1], S=S)
  l2[i] <- l_QE(c=c2[i], m=m2[i], S=S)
}
```

@@

```{r eval=TRUE, include=TRUE, results='hold'}
if(diagnostic(l1[i-1], l_hat1[i-1]) & i>2 & diagnostic(l1[i-2], l_hat1[i-2])){
  m1[i] <- m1[i-1]
  c1[i] <- c1[i-1]
  l1[i] <- l1[i-1]

  m2[i] <- m2[i-1]
  c2[i] <- c2[i-1]
  l2[i] <- l2[i-1]
}
```

@@

```{r eval=TRUE, include=TRUE, results='hold'}
i <- 4
print(paste("Iteration i =", i), quote = FALSE)
```

>>>>

```{r message=FALSE, warn=FALSE, eval=TRUE}
library(rstan)
library(cmdstanr)
options(mc.cores = parallel::detectCores())
rstan_options(auto_write = FALSE)
library(posterior)
library(bayesplot)
library(RColorBrewer)
library(ggplot2)
```

@@

where $k(\boldsymbol{x}, \boldsymbol{x}', \theta)$ is a squared exponential kernel as a function of input values $\boldsymbol{x}$ and parameters $\theta=(\alpha,\boldsymbol{\ell})$, where $\alpha$ is the marginal variance of the GP function and $\boldsymbol{\ell}=(\ell_1,\ell_2)$ the lengthscale of the kernel for dimensions 1 and 2, respectively.

@@

Matrix of 2D-predictive input points `x_pred`

```{r eval=TRUE}
x_pred <- expand.grid(x1= seq(-1,1,by=0.04), x2= seq(-1,1,by=0.04))  
str(x_pred)
```

Length of `x_pred` (`N_pred`)

```{r eval=TRUE}
N_pred <- dim(x_pred)[1]
N_pred
```

@@

Half range `S` of every input dimension

```{r eval=TRUE}
S1 <- abs((max(x_pred$x1) - min(x_pred$x1))/2)
S1
S2 <- abs((max(x_pred$x2) - min(x_pred$x2))/2)
S2
```

@@

Marginal variance `gpscale` ($\alpha$) and multivariate lengthscale `lscale` ($\boldsymbol{\ell}$)

@@

```{r eval=TRUE, results=FALSE, message=FALSE, warn=FALSE}
simuGP_mod <- cmdstanr::cmdstan_model(stan_file = "stancode_SIMU_2D.stan")
fit_SIMU <- simuGP_mod$sample(data= standata,
                                iter_warmup=1,
                                iter_sampling=1,
                                chains=1, thin=1, init=0.5,
                                fixed_param=TRUE)
fit_SIMU <- read_stan_csv(fit_SIMU$output_files())
```

@@

```{r eval=TRUE}
f_true <- extract(fit_SIMU, pars="f_true", permuted = TRUE, inc_warmup = FALSE, include = TRUE)$f_true[1,]
```

```{r eval=TRUE, echo=FALSE}
xx <- unique(x_pred[,1])

@@

## Posterior predictive distribution

```{r eval=TRUE}
f_gp <- summary(fit_GP, pars = c("f_pred"), probs = c(0.025, 0.5, 0.975), digits_summary = 4)$summary[,c(1,4,6)]
```

```{r eval=TRUE, echo=FALSE}
col1 <- brewer.pal(11,"BrBG")[11:1] 
# brk <- seq(min(f_gp[,1]),max(f_gp[,1]), length.out= 12)

filled.contour(xx, xx, matrix(f_gp[,1],length(xx),length(xx)), levels = brk, col = brewer.pal(11,"BrBG")[11:1], plot.axes= {axis(1, tick=FALSE, line=-1.1, font=5, mgp= c(2.5, 0.1, 0), las=1, cex.axis=1.5); title(xlab =expression(x[1]), mgp= c(0.8, 1, 0), cex.lab=2, las=1, family="serif"); axis(2, tick=FALSE, line=0.4, font=5, mgp= c(2.5, 0.0, 0), las=1, cex.axis=1.5); points(x_pred[vv_sample,], cex=1, lwd=1, pch=1);  abline(h= seq(-1,1,0.5), col=grey(0.7)); lines(c(-1,-1,-0.5,-0.5,0,0,0.5,0.5,1,1), c(-1.03,1.03,-1.03,1.03,-1.03,1.03,-1.03,1.03,-1.03,1.03), col=grey(0.7), type="h")}, asp= 1, frame.plot=FALSE, xlim=c(-1.03,1.03), ylim=c(-1.03,1.03)) #; grid(col=grey(0.90), lty=1)
text(-0.3,0.9,"(b)   GP", font=2, cex=1.5);
text( par("usr")[1] - 0.30, mean(c(-1,1)), labels = expression(x[2]), srt = 0, xpd = TRUE, cex=2, font=1, family="serif")
```

@@

```{r eval=TRUE}
error <- f_true - f_gp[,1]
```

```{r eval=TRUE, echo=FALSE}
col3 <- c(colorRampPalette(c("blue", "white"), bias=10)(6), colorRampPalette(c("white", "red"), bias=10)(6)[2:6])#brewer.pal(11,"RdBu")[1:11]
brk_error <- seq(min(error),max(error), length.out= 12)

@@

```{r eval=TRUE}
m_QE <- function(c,l,S) ceiling(1.75 * c / (l/S))
l_QE <- function(c,m,S) round(S * 1.75 * c / m, 3)   
```

@@

as a function of $\ell$ and $S$, where $S$ is the half range of the input domain

```{r eval=TRUE}
c_vs_l_QE <- function(l,S){
    c =  3.2*l/S
    if(c < 1.2)
      c = 1.2
  c
}
```

# Lengthscale diagnostic

The HSGP lengthscale diagnostic is defined as:

@@

```{r eval=TRUE}
diagnostic <- function(l,l_hat) l_hat + 0.01 >= l
```

@@

```{r eval=TRUE}
i <- 1
print(paste("Iteration i =", i), quote = FALSE)
```

@@

```{r eval=TRUE}
if(i==1){
  l1[i] <- 0.5
  c1[i] <- c_vs_l_QE(l=l1[i], S=S1)
  m1[i] <- m_QE(c=c1[i], l=l1[i], S=S1)

  l2[i] <- 1
  c2[i] <- c_vs_l_QE(l=l2[i], S=S2)
  m2[i] <- m_QE(c=c2[i], l=l2[i], S=S2)
}
```

@@

```{r eval=TRUE}
standata[[i]] <- list(D= D, 
                      M= c(m1[i], m2[i]),
                      M_nD= m1[i]*m2[i], 
                      indices= indices,
                      L= c(c1[i], c2[i]) * c(S1, S2), 
                      x_pred= x_pred, 
                      y_pred= y_pred, 
                      N_sample= N_sample, 
                      N_pred= N_pred, 
                      vv_sample= vv_sample
)
str(standata[[i]])
```

@@

```{r eval=TRUE}
f_hsgp <- summary(fit[[i]], pars = c("f"), probs = c(0.025, 0.5, 0.975), digits_summary = 4)$summary[,c(1,4,6)]

# rmse against data
residual <- standata[[i]]$y_pred[vv_sample] - f_hsgp[,1]
rmse[i] <- round(sqrt(mean(residual^2)), 2)

# rmse against gp
residual_gp <- f_gp[vv_sample] - f_hsgp[,1]
rmse_gp[i] <- round(sqrt(mean(residual_gp^2)), 2)

@@

```{r eval=TRUE}
l_hat1[i] <- round(summary(fit[[i]], pars = "lscale[1]", probs = c(0.05, 0.95))$summary[,1], 2)
l_hat2[i] <- round(summary(fit[[i]], pars = "lscale[2]", probs = c(0.05, 0.95))$summary[,1], 2)
```

@@

```{r eval=TRUE}
check1[i] <- diagnostic(l1[i], l_hat1[i])
check2[i] <- diagnostic(l2[i], l_hat2[i])
```

@@

  - `rmse_gp` refers to the root mean square error of the HSGP function with the GP function as the reference.

  - `rmse` refers to the root mean square error of the HSGP function with the true function as the reference.

```{r eval=TRUE}
diagnosis[[i]] <- data.frame(iter= c(i, i),
                              Dim= c("D1", "D2"),
                              l= c(l1[i], l2[i]),
                              c= c(c1[i], c2[i]),
                              m= c(m1[i], m2[i]),
                              l_hat= c(l_hat1[i], l_hat2[i]),
                              'l_hat > l' = c(check1[i], check2[i]),
                              rsme_gp = c(NA, rmse_gp[i]),
                              rsme = c(NA, rmse[i]),
                              R2 = c(NA, eR[i]),
                              elpd = c(NA, elpd[i]))
names(diagnosis[[i]]) <- c("iter", "Dim", "l", "c", "m", "l_hat", "l_hat + 0.01 > l", "rmse_gp", "rmse", "R^2", "elpd")

@@

```{r eval=TRUE}
i <- 2
print(paste("Iteration i =", i), quote = FALSE)
```

@@

```{r eval=TRUE}
if(diagnostic(l1[i-1], l_hat1[i-1]) & i==2){
  m1[i] <- m1[i-1] + 2
  c1[i] <- c_vs_l_QE(l=l_hat1[i-1], S=S1)
  l1[i] <- l_QE(c=c1[i], m=m1[i], S=S1)
}

if(diagnostic(l2[i-1], l_hat2[i-1]) & i==2){
  m2[i] <- m2[i-1] + 2
  c2[i] <- c_vs_l_QE(l=l_hat2[i-1], S=S2)
  l2[i] <- l_QE(c=c2[i], m=m2[i], S=S2)
}
```

@@

```{r eval=TRUE}
if(!diagnostic(l1[i-1], l_hat1[i-1])){
  l1[i] <- l_hat1[i-1]
  c1[i] <- c_vs_l_QE(l=l1[i], S=S1)
  m1[i] <- m_QE(c=c1[i], l=l1[i], S=S1)
}

if(!diagnostic(l2[i-1], l_hat2[i-1])){
  l2[i] <- l_hat2[i-1]
  c2[i] <- c_vs_l_QE(l=l2[i], S=S2)
  m2[i] <- m_QE(c=c2[i], l=l2[i], S=S2)
}
```

@@

```{r eval=TRUE}
standata[[i]] <- list(D= D, 
                      M= c(m1[i], m2[i]),
                      M_nD= m1[i]*m2[i], 
                      indices= indices,
                      L= c(c1[i], c2[i]) * c(S1, S2), 
                      x_pred= x_pred, 
                      y_pred= y_pred, 
                      N_sample= N_sample, 
                      N_pred= N_pred, 
                      vv_sample= vv_sample
)
str(standata[[i]])
```

@@

```{r eval=TRUE}
l_hat1[i] <- round(summary(fit[[i]], pars = "lscale[1]", probs = c(0.05, 0.95))$summary[,1], 3)
l_hat2[i] <- round(summary(fit[[i]], pars = "lscale[2]", probs = c(0.05, 0.95))$summary[,1], 3)
```

@@

```{r eval=TRUE}
i <- 3
print(paste("Iteration i =", i), quote = FALSE)
```

@@

```{r eval=TRUE}
if(diagnostic(l1[i-1], l_hat1[i-1]) & i>2 & !diagnostic(l1[i-2], l_hat1[i-2])){
  m1[i] <- m1[i-1] + 2
  c1[i] <- c_vs_l_QE(l=l_hat1[i-1], S=S1)
  l1[i] <- l_QE(c=c1[i], m=m1[i], S=S1)
}

if(diagnostic(l2[i-1], l_hat2[i-1]) & i>2 & !diagnostic(l2[i-2], l_hat2[i-2])){  
  m2[i] <- m2[i-1] + 2
  c2[i] <- c_vs_l_QE(l=l_hat2[i-1], S=S2)
  l2[i] <- l_QE(c=c2[i], m=m2[i], S=S2)
}
```

@@

```{r eval=TRUE}
if(diagnostic(l1[i-1], l_hat1[i-1]) & i>2 & diagnostic(l1[i-2], l_hat1[i-2])){
  m1[i] <- m1[i-1]
  c1[i] <- c1[i-1]
  l1[i] <- l1[i-1]
}

if(diagnostic(l2[i-1], l_hat2[i-1]) & i>2 & diagnostic(l2[i-2], l_hat2[i-2])){  
  m2[i] <- m2[i-1]
  c2[i] <- c2[i-1]
  l2[i] <- l2[i-1]
}
```

@@

```{r eval=TRUE}
i <- 4
print(paste("Iteration i =", i), quote = FALSE)
```
